Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-12-15 paper/manuscript_Adam2.tex
added second round oof Adam's comments

But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis,klink2017parallel}.
We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as \emph{shifts} in a site's amino-acid preferences.
Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.
Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.

@@

A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
(Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters.)
At sites of known functional importance, these preferences are usually consistent with prior knowledge.
For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}).
Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.

@@

Interestingly, immune selection also offers a plausible explanation for the sites that evolve \emph{slower} than expected.
In addition to escaping immunity via substitutions at antibody-binding footprints, Env is notorious for employing a range of more general strategies to reduce its susceptibility to antibodies.
These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments -- and indeed, such sites evolve more slowly than expected from our experiments.
For instance, sites of N-linked glycans change slower in nature than our experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}), as do sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.

@@

But these maps come with a potentially serious caveat: each one is measured for just a single Env variant.
The major question that our study aimed to answer is whether the maps are still useful for evolutionary questions, or whether Env's amino-acid preferences shift so rapidly that each map only applies to the specific HIV strain for which it was measured.
This question is reminiscent of one that was grappled with in the early days of protein crystallography, when it first became possible to build maps of a protein's structure.
Because it was not (and is still not) possible to crystallize every variant of a protein, it was necessary to determine whether protein structures could be usefully generalized among homologs.
Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter aspects of Env's conformation and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate}, all characterized Env structures remain highly similar in most respects~\citep{julien2015design,pugach2015native,stewart2016trimeric,verkerke2016epitope,gristick2017x}.

@@

Going forward, it will be interesting to determine how the results of experiments such as ours are affected by other defined modifications in viral genotype or selective pressure.
A challenge of studying HIV is that almost everything must be simplified to make it tractable.
Our studies are no different -- after all, we are measuring viral growth in a T-cell line in the lab, not in infected humans.
But our experiments provide highly quantitative data that is amenable to rigorous analysis.
Here we have shown how these data can be compared between Envs to identify sites where mutational tolerance shifts with viral genotype, or between experiments and nature to identify sites under different pressure in the lab and in humans.
Future experiments that modulate selection pressures in other relevant ways should provide further insight into the forces that drive and constrain HIV's evolution.

>>>>

But interpreting these studies in the context of Env evolution requires addressing a fundamental question: How informative are mutational studies of a single protein variant about constraints on long-term evolution?
During protein evolution, substitutions at one site can change the effect of mutations at other sites~\citep{natarajan2013epistasis,gong2013stability,harms2014historical,podgornaia2015pervasive,starr2016epistasis,klink2017parallel}.
We will follow the nomenclature of \citet{pollock2012amino} to refer to these changes in mutational effects as \emph{shifts} in a site's amino-acid preferences.
Such shifts can accumulate as substitutions become entrenched via epistatic interactions with subsequent changes~\citep{starr2017pervasive,pollock2012amino,shah2015contingency,bazykin2015changing} -- although the magnitude of these shifts is usually limited~\citep{doud2015site,chan2017correlation,ashenberg2013mutational,risso2014mutational}.

Given that the Envs of circulating HIV strains represent a vast collection of homologs that often differ at $>$100 residues, such shifts could make the outcome of any study highly dependent on the Env used.
The extent to which this is actually the case is unclear, since the few protein-wide studies of mutational shifts have examined proteins that are structurally far simpler than Env, which forms a large heavily glycosylated heterotrimeric complex that transitions through multiple conformational states~\citep{munro2014conformational,ozorowski2017open}.

@@

A more qualitative way to assess if the deep mutational scanning authentically describes selection on Env function is to visually compare the measurements with existing knowledge.
Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} show the across-replicate average of the amino-acid preferences for each Env after re-scaling by the stringency parameters in Table~\ref{tab:phydms}.
(Note that throughout the rest of this paper, we use preferences re-scaled by these stringency parameters.)
At sites of known functional importance, these preferences are usually consistent with prior knowledge.
For instance, residues T257, D368, E370, W427, and D457 are important for Env binding to CD4~\citep{olshevsky1990identification}, and all these amino acids are highly preferred in our deep mutational scanning (Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}).
Likewise, Env has 10 disulfide bonds (linking sites 54-74, 119-205, 126-196, 131-157, 218-247, 228-239, 296-331, 378-445, 385-418, and 598-604), most of which are important for function~\citep{van2008only} -- and the cysteines at these sites are highly preferred in our deep mutational scanning.
The deep mutational scanning is also consistent with prior knowledge about sites that are tolerant of mutations.
For instance, Env has five variable loops that mostly evolve under weak constraint in nature~\citep{starcich1986identification,zolla2010structure} -- and most sites in these loops are mutationally tolerant in our deep mutational scanning (see sites indicated by gray overlay bars in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs}, such as 132 to 195).
Intringuingly, we find very little selection to maintain glycosylation in our cell culutre experiments.
Of the 21 N-linked glycosylations sites shared between BG505 and BF52, only 4 are under strong selection to maintain the glycan in our experiments while they are largely conserved in nature (Figure 9S2).
This highlights the role that these glycans play in immune selection, which is absent in our experiments, and agree with/adds to previous literature (cite). 
It is beyond the scope of this paper to catalog associations between our measurements and all other prior mutational studies of Env, but we anticipate that the complete maps of mutational effects in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} will be useful for future sequence-structure-function studies.

@@

Interestingly, immune selection also offers a plausible explanation for the sites that evolve \emph{slower} than expected.
In addition to escaping immunity via substitutions at antibody-binding footprints, Env is notorious for employing a range of more general strategies to reduce its susceptibility to antibodies.
These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments -- and indeed, such sites evolve more slowly than expected from our experiments.
For instance, many sites of N-linked glycans change slower in nature than our would be expected based on our experimental measurements of cell culture replication (Figure~\ref{fig:divsel}C and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}). Similarly, a network of sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D) also evolve slower in nature than expected.
Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.

@@

But these maps come with a potentially serious caveat: each one is measured for just a single Env variant.
The major question that our study aimed to answer is whether the maps are still useful for evolutionary questions, or whether Env's amino-acid preferences shift so rapidly that each map only applies to the specific HIV strain for which it was measured.
This question is reminiscent of one that was grappled with in the early days of protein crystallography, when it first became possible to build maps of a protein's structure.
Because it was not (and is still not) possible to crystallize every variant of a protein, it was necessary to determine whether protein structures could be usefully generalized among homologs.
Fortunately for the utility of structural biology, it soon became apparent that closely homologous proteins have similar structures~\citep{chothia1986relation,sander1991database}.
This rough generalizability of protein structures holds even for a protein as conformationally complex as Env -- for although there are many examples of mutations that alter aspects of Env's conformation and dynamics~\citep{kwong2000structures,white2010molecular,almond2010structural,davenport2013isolate},  SOSIP trimer structures from diverse Env strains remain highly similar in most respects~\citep{julien2015design,pugach2015native,stewart2016trimeric,verkerke2016epitope,gristick2017x}.

@@

These experiments provide highly quantitative datasets on the functional tolerance of Env that is amenable to rigorous functional and evolutionary analyses.
Here we have shown how these data can be compared between Envs to identify sites where mutational tolerance shifts with viral genotype, or between experiments and nature to identify sites under different pressure in the lab and in humans.
Future experiments that modulate selection pressures in other relevant ways should provide further insight into the forces that drive and constrain HIV's evolution.
